Regionally Additive Models: Explainable-by-design models minimizing feature interactions

### 2023-06-21 report/report.tex
update

\begin{document}
\maketitle

@@

\begin{figure}[htbp]
    \centering
    \begin{subfigure}{0.32\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/global_GAM}
        \caption{\(f_2(x_2)\)}
        \label{subfig:global_gam}
    \end{subfigure}
    \begin{subfigure}{0.32\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/regional_gam_subreg_1}
        \caption{\(f_2(x_2) \when{x_1 > 0 \text{ and } x_3 = 1}\)}
        \label{subfig:regional_gam_1}
    \end{subfigure}
    \begin{subfigure}{0.32\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/regional_gam_subreg_2}
        \caption{\(f_2(x_2) \when{x_1 \leq 0 \text{ or } x_3 \neq 1}\)}
        \label{subfig:regional_gam_2}
    \end{subfigure}
    \caption{Caption for the entire figure}
    \label{fig:ram_example}
\end{figure}

@@

In the initial step of the pipeline, we fit a black-box function \(f(\cdot)\) to the training set
\(\mathcal{D} = \{(\xb^i, y^i)\}_{i=1}^N\) to accurately learn the underlying mapping \(f(\cdot) : \Xcal \rightarrow \Ycal\).
While any black-box function can theoretically be employed in this stage,
for utilizing the DALE approximation, as we will show in the next step,
it is necessary to select a differentiable function.
Recent advancements have demonstrated that differentiable Deep Learning models,
specifically designed for tabular data~\citep{arik2021tabnet}, are capable of achieving state-of-the-art performance,
making them a suitable choice for this step.

@@

For minimizing Eq.~\eqref{eq:SE}, we develop a tree-based algorithm, similar to the one proposed by~\citep{herbinger2023decomposing}.
The backbone of the algorithm is described in Algorithm~\ref{alg:subregion_detection}.
For each feature $s$, we search for the $T$ optimal splits.
To better understand the algorithm, let's take the toy example of Section~\ref{sec:motivation}.
The subregions are searched independently for each feature $s \in \{1, \ldots, D\}$.
For feture $s=2$, we start by searching for the optimal split for the first level of the tree.
The candidate features for defining the first level split are $x_1$ and $x_3$.
The algorithm~\ref{alg:single_feature_subregion} is responsible to find the optimal first level split.
First, it defines the candidate split points for each feature.
Since $x_1$ is a continuous feature, the candidate split points are a linearly spaced grid of $p$ points in the range $[-1, 1]$,
where $P$ is a hyperparameter of the algorithm, set to 10 in the experiments.
Therefore, the candidate split points are $p = \{-1, -0.8, -0.6, \ldots, 0.8, 1\}$ and
the corrsponding subregions are $\mathcal{R}_{21} = \{ \mathbb{R}^2 : x_1 \leq p \}$ and $\mathcal{R}_{22} = \{ \mathbb{R}^2 : x_1 > p \}$.
Since $x_3$ is categorical, the candidate split points are the unique values of $x_3$, i.e. $\{0, 1\}$
and the corresponding subregions are $\mathcal{R}_{21} = \{ \mathbb{R}^2 : x_3 = 0 \}$ and $\mathcal{R}_{22} = \{ \mathbb{R}^2 : x_3 = 1 \}$.
For each candidate split point, the algorithm computes the corresponding subregions and the corresponding $\mathcal{H}_{\mathcal{R}_{st}}$.
and selects the split point that minimizes the weighted level of interactions, as defined by Algorithm~\ref{alg:get_interaction}.
Suppose that the optimal first-level subregions are $\mathcal{R}_{21} = \{ \mathbb{R}^2 : x_3 = 0 \}$ and $\mathcal{R}_{22} = \{ \mathbb{R}^2 : x_3 \neq 0 \}$.
The algorithm then proceeds to the second level of the tree.
For the second level, the first split is fixed, i.e. the are two datasets split according to the first-level split and the algorithm searches for the optimal second-level split.
The candidate features for defining the second level split are $x_3$.

The output of algorithm~\ref{alg:subregion_detection} is a set of $T$ splits per feature $s$.
It is imporant to note, that since not all splits lead to an important decrease in the level of interactions,
we post-process the splits by pruning all splits that do not decrease the loss function by more than a threshold $\epsilon$.
The final set of splits is denoted as $\{\mathcal{R}_{st}\}_{t=1}^{T_s}$.

@@

Algorithm~\ref{alg:subregion_detection} has a computational complexity of $\mathcal{O}(D \cdot T)$.
and Algorithm~\ref{alg:single_feature_subregion} has a computational complexity of $\mathcal{O}(D-1 \cdot P \cdot N)$
where $D$ is the number of feature, $N$ is the number of samples, $T$ is the maximum depth of the tree and $P$ is the number of query postitions.
The computational complexity of the proposed method is therefore $\mathcal{O}(D \cdot (D-1) \cdot T \cdot P \cdot N)$.
However, in practice, the number of query positions $P$ and the maximum depth $T$ are small numbers.
Therefore, the computational complexity of the proposed method is $\mathcal{O}(D^2 \cdot N)$.

\subsection{Fitting the GAMs}
\label{subsec:fitting_gams}

\subsubsection{Objective}

\subsection{Discussion}

Recently, a number of methods have been proposed to extend traditional GAMs and make them more expressive.
The majority of the ideas follow one of the following research directions;
The first one targets on representing the main components of a GAM $\{ f_i(x_i) \}$ with novel models.
For example, \citet{agarwal2021neural} who used an end-to-end neural network for learning the main components.
The second one focuses on extending the GAMs to model interactions between the features.
For example, \citet{lou2013accurate} proposed Explainable Boosting Machines (EBMs) which are a generalized additive model with pairwise interaction terms.
It is worth noting that the proposed method is orthogonal to the aforementioned research directions.
As we will show in the experiments, the proposed method can be used in conjunction with any of the
aforementioned methods to improve the accuracy of the resulting model, while maintaining the interpretability of the model.

@@

\begin{table}[htbp]
  \centering
  \caption{Result Comparison}
  \label{tab:sample}
  \begin{tabular}{l|c|cccc}
      \hline
      & \textbf{Black-box} & \multicolumn{4}{c}{\textbf{x-by-design}} \\
      \hline
      \hline
      & all orders & \multicolumn{2}{c}{1\textsuperscript{st} order} & \multicolumn{2}{c}{2\textsuperscript{nd} order} \\
      \hline
      \hline
      & \textbf{DNN} & \textbf{GAM} & \textbf{RAM} & \textbf{GAM}$^2$ & \textbf{RAM}$^2$ \\
      \hline
      Bike Sharing      & 0.254 & 0.549 & 0.430 & 0.298 & 0.278 \\
      \hline
      California Housing      & 0.369 & 0.600 & 0.535 & 0.554 & 0.514 \\
  \end{tabular}
\end{table}

@@

@@

The California Housing dataset consists of approximately $20,000$ of housing blocks situated in California.
Each housing block is described by eight numerical features, namely,
$X_{\mathtt{latitude}}$, $X_{\mathtt{longitude}}$, $X_{\mathtt{median\_age}}$, $X_{\mathtt{total\_rooms}}$, $X_{\mathtt{total\_bedrooms}}$, $X_{\mathtt{population}}$, $X_{\mathtt{households}}$, and $X_{\mathtt{median\_income}}$.
The target variable, $Y_{\mathtt{value}}$, is the median house value in dollars for each block.
The target value ranges in the interval \([15, 500] \cdot 10^3\), with a mean value of
\(\mu_Y \approx 201 \cdot 10^3 \) and a standard deviation of \(\sigma_Y \approx 110 \cdot 10^3\).

As a black-box model, we train for $45$ epochs a fully-connected Neural Network with 6 hidden layers,
using the Adam optimizer with a learning rate of $0.001$.
The model achieves a root mean square error (RMSE) of about \(40\)K dollars on the test set.
Subsequently, we perform subregion extraction by searching for splits up to a maximum depth of \(T=3\).
After the postprocessing step, we discover that several splits significantly reduce the level of interactions,
resulting in an expanded input space consisting of \(16\) features, as we show in table~\ref{tab:california_housing_subregions}.
Out of them, we randomly select and illustrate in Figure~\ref{fig:california_housing} the effect of the feature $X_{\mathtt{longitude}}$.
As we observe, for the house blocks located in the southern part of California ($X_{\mathtt{latitude}} \leq 34.9$),
the house value decreases in an almost linear fashion as we move eastward ($X_{\mathtt{longitude}}$ increases).
In contrast, for the house blocks located in the northern part of California ($X_{\mathtt{latitude}} > 34.9$),
the house value performs a rapid (non-linear) decrease as we move eastward ($X_{\mathtt{longitude}}$ increases).
We also observe that although the EBM fitted to each subregion captures the general trend,
it does not align perfectly with the regional effect.
As in the Bike-Sharing Example, the RMSE of the RAM model, i.e. \( 0.53 \cdot 110 \cdot 10^3 \approx 58.3 \cdot 10^3\) dollars on the test set,
is lower than the one of the GAM model, i.e.\( 0.6 \cdot 110 \cdot 10^3 \approx 66 \cdot 10^3\) dollars.
These results indicate that the RAM model provides superior predictions compared to the GAM model.
The same conclusion holds is when comparing the RAM$^2$ and the GAM$^2$ models.
\begin{table}[htbp]
  \centering
  \caption{California Housing: Subregions Detected by RAM}
  \label{tab:california_housing_subregions}
  \begin{tabular}{c|c|}
      Feature & Subregions \\
      \hline
      \multirow{2}{*}{$X_{\mathtt{longitude}}$} & $X_{\mathtt{longitude}} \when{X_{\mathtt{latitude}} \leq 34.9}$ \\
      & $X_{\mathtt{longitude}} \when{X_{\mathtt{latitude}} > 34.9}$ \\
      \hline
      \multirow{2}{*}{$X_{\mathtt{latitude}}$} & $X_{\mathtt{latitude}} \when{X_{\mathtt{longitude}} \leq -120.31}$ \\
      & $X_{\mathtt{latitude}} \when{X_{\mathtt{longitude}} > -120.31}$ \\
      \hline
      \multirow{2}{*}{$X_{\mathtt{total\_rooms}}$} & $X_{\mathtt{total\_rooms}} \when{X_{\mathtt{total\_bedrooms}} \leq 449.37}$ \\
        & $X_{\mathtt{total\_rooms}} \when{X_{\mathtt{total\_bedrooms}} > 449.37}$ \\
      \hline
      \multirow{4}{*}{$X_{\mathtt{total\_bedrooms}}$} & $X_{\mathtt{total\_bedrooms}} \when{X_{\mathtt{households}} \leq 411} \when{X_{\mathtt{total\_bedrooms}} \leq 647}$ \\
        & $X_{\mathtt{total\_bedrooms}} \when{X_{\mathtt{households}} \leq 411} \when{X_{\mathtt{total\_bedrooms}} > 647}$ \\
        & $X_{\mathtt{total\_bedrooms}} \when{X_{\mathtt{households}} > 411} \when{X_{\mathtt{total\_bedrooms}} \leq 647}$ \\
        & $X_{\mathtt{total\_bedrooms}} \when{X_{\mathtt{households}} > 411} \when{X_{\mathtt{total\_bedrooms}} > 647}$ \\
      \hline
      \multirow{2}{*}{$X_{\mathtt{population}}$} & $X_{\mathtt{population}} \when{X_{\mathtt{households}} \leq 411.5}$ \\
      & $X_{\mathtt{population}} \when{X_{\mathtt{households}} > 411.5}$ \\
      \hline
      \multirow{2}{*}{$X_{\mathtt{households}}$} & $X_{\mathtt{households}} \when{X_{\mathtt{total\_bedrooms}} \leq 630.57}$ \\
        & $X_{\mathtt{households}} \when{X_{\mathtt{total\_bedrooms}} > 630.57}$ \\
  \end{tabular}
\end{table}

\begin{figure}[htbp]
    \centering
    \begin{subfigure}{0.32\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/california_gam}
        \caption{\(f(X_{\mathtt{longitude}})\)}
        \label{subfig:california_gam}
    \end{subfigure}
    \begin{subfigure}{0.32\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/california_ram_1}
        \caption{\(f(X_{\mathtt{longitude}}) \when{X_{\mathtt{latitude}} \leq 34.89}\)}
        \label{subfig:california_ram_1}
    \end{subfigure}
    \begin{subfigure}{0.32\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/california_ram_2}
        \caption{\(f(X_{\mathtt{longitude}}) \when{X_{\mathtt{latitude}} > 34.89}\)}
        \label{subfig:california_ram_2}
    \end{subfigure}
    \caption{Caption for the entire figure}
    \label{fig:california_housing}
\end{figure}

@@

\section{Appendix}

\subsection{Regional DALE formulation}
\label{sec:regional_dale}

For this reason we define the bin-effect, the bin-deviation and the aggregated deviation on a subregion \(\mathcal{R}_{st}\) as:

\begin{equation}
    \label{eq:bin_effect_subregion}
    \hat{\mu}_{\mathcal{R}_{st}}(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k \cap \mathcal{R}_{st}|} \sum_{i:\mathbf{x}^i \in \mathcal{S}_k \cap \mathcal{R}_{st} } \dfdx(\mathbf{x}^i)
\end{equation}

\begin{equation}
  \label{eq:bin_deviation_subregion}
  \hat{\sigma}_{\mathcal{R}_{st}}^2(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k \cap \mathcal{R}_{st}| - 1}
\sum_{i:\mathbf{x}^i \in \mathcal{S}_k \cap \mathcal{R}_{st} } \left ( \dfdx(\mathbf{x}^i) -
    \hat{\mu}(z_{k-1}, z_k) \right )^2
\end{equation}

\begin{equation}
  \label{eq:SE}
  \mathcal{L}_{\mathcal{R}_{sk}} = \sum_{k=1}^K (z_k - z_{k-1})^2 \hat{\sigma}_{\mathcal{R}_{st}}^2(z_{k-1}, z_k)
\end{equation}

\subsection{Algorithmic Details of Subregion Detection}

\begin{algorithm}
\caption{SplitDataset}
\label{alg:split_dataset}
\SetAlgoLined
\SetKwInOut{Input}{Input}
\SetKwInOut{Output}{Output}
\Input{X\_list, J\_list, c, val, is\_cat}
\Output{X\_split, J\_split}
X\_split $\gets$ []\;
J\_split $\gets$ []\;
\For{i = 1 to len(X\_list)}{
    X= X\_list[i]\;
    J= J\_list[i]\;
    \uIf{is\_cat}{
        ind\_1 $\gets$ X[:, c] $=$ val\;
        ind\_2 $\gets$ X[:, c] $\neq$ val\;
    }
    \Else{
        ind\_1 $\gets$ X[:, c] $\leq$ val\;
        ind\_2 $\gets$ X[:, c] $>$ val\;
    }
    Append X[ind\_1], X[ind\_2] to X\_split\;
    Append J[ind\_1], J[ind\_2] to J\_split\;
}
\Return X\_split, J\_split
\end{algorithm}

>>>>

\begin{document}
    \def\MakeUppercaseUnsupportedInPdfStrings{\scshape}
\maketitle

@@

\begin{figure}[htbp]
    \centering
    \begin{subfigure}{0.32\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/global_GAM}
        \caption{\(f_2(x_2)\)}
        \label{subfig:global_gam}
    \end{subfigure}
    \begin{subfigure}{0.32\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/regional_gam_subreg_1}
        \caption{\(f_2(x_2) \when{x_1 > 0 \text{ and } x_3 = 1}\)}
        \label{subfig:regional_gam_1}
    \end{subfigure}
    \begin{subfigure}{0.32\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/regional_gam_subreg_2}
        \caption{\(f_2(x_2) \when{x_1 \leq 0 \text{ or } x_3 \neq 1}\)}
        \label{subfig:regional_gam_2}
    \end{subfigure}
    \caption{The left image showcases the global GAM which erroneously learns an approximation of \(f(\xb) \approx 2x_2\).
    In contrast, the middle and right images demonstrate the RAM's ability to identify two distinct subregions where \(f\) exhibits regional additivity.
    By fitting a GAM to each subregion, the RAM accurately captures the true function $f$ while retaining interpretability.}
    \label{fig:ram_example}
\end{figure}

@@

In the initial step of the pipeline, we fit a black-box function \(f(\cdot)\) to the training set
\(\mathcal{D} = \{(\xb^i, y^i)\}_{i=1}^N\) to accurately learn the underlying mapping \(f(\cdot) : \Xcal \rightarrow \Ycal\).
While any black-box function can theoretically be employed in this stage,
for utilizing the DALE approximation, as we will show in the next step,
it is necessary to select a differentiable function.
Recent advancements have demonstrated that differentiable Deep Learning models,
specifically designed for tabular data~\citep{arik2021tabnet}, are capable of achieving state-of-the-art performance,
making them a suitable choice for this step.
% Driver, teacher, guide

@@

To minimize the objective of Eq.~\eqref{eq:SE}, we have developed a tree-based algorithm based on the approach proposed
by~\citep{herbinger2023decomposing}.
The core of the algorithm is outlined in Algorithm~\ref{alg:subregion_detection}.
We search for subregions independently for each feature $s \in \{1, \ldots, D\}$.
and we select the $T$ optimal splits, in a greedy manner.
To better comprehend the algorithm, let's consider the illustrative example of Section~\ref{sec:motivation}.
For feature $s=2$, the algorithm starts with the first level of the tree, with $/s = \{1, 3\}$ as candidate split-features.
Algorithm~\ref{alg:single_feature_subregion} identifies the optimal first-level split, with the following procedure.
Initially, it determines the candidate split points for each feature.

Since $x_1$ is a continuous feature, are generated as a linearly spaced grid of $P$ points within the range of the feature, i.e. $[-1, 1]$,
where $P$ is a hyperparameter of the algorithm, set to 10 in the experiments.
Consequently, the candidate split points $p \in \{-1, -0.8, -0.6, \ldots, 0.8, 1\}$ define the candidate subregions
$\mathcal{R}_{21} = \{ \mathbb{R}^2 : x_1 \leq p \}$ and $\mathcal{R}_{22} = \{ \mathbb{R}^2 : x_1 > p \}$.
As for $x_3$, being a categorical feature, the candidate split points are its unique values, i.e., $\{0, 1\}$,
and the corresponding subregions are $\mathcal{R}_{21} = \{ \mathbb{R}^2 : x_3 = 0 \}$ and $\mathcal{R}_{22} = \{ \mathbb{R}^2 : x_3 \neq 0 \}$.

For each candidate position, the algorithm splits the initial dataset into two subsets,
$[\mathcal{D}_1, \mathcal{D}_2]$ one for each subregion and computes the level of interaction
as the weighted sum of the interactions of the two subregions, as defined by Algorithm~\ref{alg:get_interaction}.
After the iteration over all points and all candidate positions,
the split point that minimizes the weighted level of interactions is selected.
In the illustrative example, the optimal first-level subregions are
$\mathcal{R}_{21} = \{ \mathbb{R}^2 : x_3 = 0 \}$ and $\mathcal{R}_{22} = \{ \mathbb{R}^2 : x_3 \neq 0 \}$.
The algorithm then proceeds to the second level of the tree, where the only candidate feature is $x_3$.
In this step, the first split is considered fixed, i.e., the algorithm uses a list of two datasets,
and searches in an analogous manner for the optimal second-level split that will be applied on top of the first split,
leading to four subregions.
The algorithm proceeds in a similar manner, until it reaches the maximum depth $T$.

The output of algorithm~\ref{alg:subregion_detection} is a collection of $T$ splits for each feature $s$.
It is important to note that not all splits lead to a significant decrease in the level of interactions.
To address this, we post-process the splits by pruning any splits that do not result in a decrease of the
weighted interaction greater than a threshold value $\epsilon$ (set to $20\%$ drop in the experiments).
The pruned set of splits is denoted as $\{\mathcal{R}_{st}\}_{t=1}^{T_s}$.

@@

Algorithm~\ref{alg:subregion_detection} has a computational complexity of $\mathcal{O}(D \cdot T)$
as it iterates over all features and levels of the tree.
Algorithm~\ref{alg:single_feature_subregion} has a computational complexity of $\mathcal{O}(D \cdot P \cdot N)$
as it iterates over all features, query positions, and performs indexing operations on the data (SplitDataset and GetInteraction).
The use of DALE eliminates the need to compute the Jacobian matrix for each split,
resulting in computational complexity independent of the model complexity.
Considering the entire algorithm, the computational complexity is $\mathcal{O}(D \cdot (D-1) \cdot T \cdot P \cdot N)$.
However, in practice, $P$ and $T$ are small numbers.
Therefore, the computational complexity of the proposed method simplifies to $\mathcal{O}(D^2 \cdot N)$,
making it suitable for large datasets, heavy models, and reasonably high-dimensional data.
In practice, for a typical scenario with $N=10 \times 10^3$ samples and $D=20$ features, the proposed method takes less than $10$ seconds to run.

\subsection{Third step: Fit a GAM in each subregion}
\label{subsec:fitting_gams}

Once the subregions are detected, any Generalized Additive Model (GAM) family can be fitted to the augmented input space $\mathcal{X}^{\mathtt{RAM}}$.
Recently, several methods have been proposed to extend GAMs and enhance their expressiveness.
These methods can be categorized into two main research directions.
The first direction targets on representing the main components of a GAM $\{ f_i(x_i) \}$ using novel models.
For example,~\citep{agarwal2021neural} introduced an approach that employs an end-to-end neural network to learn the main components.
The second direction aims to extend GAMs to model feature interactions.
Examples of such extensions include Explainable Boosting Machines (EBMs)~\citep{lou2013accurate} or
Node-GAMs~\citep{chang2021node}.
These models are generalized additive models that incorporate pairwise interaction terms.
It is worth noticing, that the RAM framework and can be used on top of both these research directions
to further enhance the expressiveness of the models while maintaining their interpretability.
In our experiments, we use the Explainable Boosting Machines (EBMs).

@@

@@

\begin{table}[htbp]
  \centering
  \caption{The table compares RMSE values of different models on two datasets: Bike-Sharing and California Housing.
  The models include DNN, GAM, RAM, GAM², and RAM² (representing 2nd order interactions).
  Lower values indicate better performance.
  RAM consistently outperforms GAM and approaches DNN performance.}
  \label{tab:sample}
  \begin{tabular}{l|c|cccc}
      \hline
      & \textbf{Black-box} & \multicolumn{4}{c}{\textbf{x-by-design}} \\
      \hline
      \hline
      & all orders & \multicolumn{2}{c}{1\textsuperscript{st} order} & \multicolumn{2}{c}{2\textsuperscript{nd} order} \\
      \hline
      \hline
      & \textbf{DNN} & \textbf{GAM} & \textbf{RAM} & \textbf{GAM}$^2$ & \textbf{RAM}$^2$ \\
      \hline
      Bike Sharing      & 0.254 & 0.549 & 0.430 & 0.298 & 0.278 \\
      \hline
      California Housing      & 0.369 & 0.600 & 0.535 & 0.554 & 0.514 \\
  \end{tabular}
\end{table}

@@

The California Housing dataset consists of approximately $20,000$ of housing blocks situated in California.
Each housing block is described by eight numerical features, namely,
$X_{\mathtt{lat}}$, $X_{\mathtt{long}}$, $X_{\mathtt{median\_age}}$, $X_{\mathtt{total\_rooms}}$, $X_{\mathtt{total\_bedrooms}}$, $X_{\mathtt{population}}$, $X_{\mathtt{households}}$, and $X_{\mathtt{median\_income}}$.
The target variable, $Y_{\mathtt{value}}$, is the median house value in dollars for each block.
The target value ranges in the interval \([15, 500] \cdot 10^3\), with a mean value of
\(\mu_Y \approx 201 \cdot 10^3 \) and a standard deviation of \(\sigma_Y \approx 110 \cdot 10^3\).

As a black-box model, we train for $45$ epochs a fully-connected Neural Network with 6 hidden layers,
using the Adam optimizer with a learning rate of $0.001$.
The model achieves a root mean square error (RMSE) of about \(40\)K dollars on the test set.
Subsequently, we perform subregion extraction by searching for splits up to a maximum depth of \(T=3\).
After the postprocessing step, we discover that several splits significantly reduce the level of interactions,
resulting in an expanded input space consisting of \(16\) features, as we show in table~\ref{tab:california_housing_subregions}.
Out of them, we randomly select and illustrate in Figure~\ref{fig:california_housing} the effect of the feature $X_{\mathtt{long}}$.
As we observe, for the house blocks located in the southern part of California ($X_{\mathtt{lat}} \leq 34.9$),
the house value decreases in an almost linear fashion as we move eastward ($X_{\mathtt{long}}$ increases).
In contrast, for the house blocks located in the northern part of California ($X_{\mathtt{lat}} > 34.9$),
the house value performs a rapid (non-linear) decrease as we move eastward ($X_{\mathtt{long}}$ increases).
We also observe that although the EBM fitted to each subregion captures the general trend,
it does not align perfectly with the regional effect.
As in the Bike-Sharing Example, the RMSE of the RAM model, i.e. \( 0.53 \cdot 110 \cdot 10^3 \approx 58.3 \cdot 10^3\) dollars on the test set,
is lower than the one of the GAM model, i.e.\( 0.6 \cdot 110 \cdot 10^3 \approx 66 \cdot 10^3\) dollars.
These results indicate that the RAM model provides superior predictions compared to the GAM model.
The same conclusion holds is when comparing the RAM$^2$ and the GAM$^2$ models.
\begin{table}[htbp]
  \centering
  \caption{California Housing: Subregions Detected by RAM}
  \label{tab:california_housing_subregions}
  \begin{tabular}{c|c}
      Feature & Subregions \\
      \hline
      \multirow{2}{*}{$X_{\mathtt{long}}$} & $X_{\mathtt{long}} \when{X_{\mathtt{lat}} \leq 34.9}$ \\
      & $X_{\mathtt{long}} \when{X_{\mathtt{lat}} > 34.9}$ \\
      \hline
      \multirow{2}{*}{$X_{\mathtt{lat}}$} & $X_{\mathtt{lat}} \when{X_{\mathtt{long}} \leq -120.31}$ \\
      & $X_{\mathtt{lat}} \when{X_{\mathtt{long}} > -120.31}$ \\
      \hline
      \multirow{2}{*}{$X_{\mathtt{total\_rooms}}$} & $X_{\mathtt{total\_rooms}} \when{X_{\mathtt{total\_bedrooms}} \leq 449.37}$ \\
        & $X_{\mathtt{total\_rooms}} \when{X_{\mathtt{total\_bedrooms}} > 449.37}$ \\
      \hline
      \multirow{4}{*}{$X_{\mathtt{total\_bedrooms}}$} & $X_{\mathtt{total\_bedrooms}} \when{X_{\mathtt{households}} \leq 411} \when{X_{\mathtt{total\_bedrooms}} \leq 647}$ \\
        & $X_{\mathtt{total\_bedrooms}} \when{X_{\mathtt{households}} \leq 411} \when{X_{\mathtt{total\_bedrooms}} > 647}$ \\
        & $X_{\mathtt{total\_bedrooms}} \when{X_{\mathtt{households}} > 411} \when{X_{\mathtt{total\_bedrooms}} \leq 647}$ \\
        & $X_{\mathtt{total\_bedrooms}} \when{X_{\mathtt{households}} > 411} \when{X_{\mathtt{total\_bedrooms}} > 647}$ \\
      \hline
      \multirow{2}{*}{$X_{\mathtt{population}}$} & $X_{\mathtt{population}} \when{X_{\mathtt{households}} \leq 411.5}$ \\
      & $X_{\mathtt{population}} \when{X_{\mathtt{households}} > 411.5}$ \\
      \hline
      \multirow{2}{*}{$X_{\mathtt{households}}$} & $X_{\mathtt{households}} \when{X_{\mathtt{total\_bedrooms}} \leq 630.57}$ \\
        & $X_{\mathtt{households}} \when{X_{\mathtt{total\_bedrooms}} > 630.57}$ \\
  \end{tabular}
\end{table}

\begin{figure}[htbp]
    \centering
    \begin{subfigure}{0.32\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/california_gam}
        \caption{\(f(X_{\mathtt{long}})\)}
        \label{subfig:california_gam}
    \end{subfigure}
    \begin{subfigure}{0.32\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/california_ram_1}
        \caption{\(f(X_{\mathtt{long}}) \when{X_{\mathtt{lat}} \leq 34.89}\)}
        \label{subfig:california_ram_1}
    \end{subfigure}
    \begin{subfigure}{0.32\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/california_ram_2}
        \caption{\(f(X_{\mathtt{long}}) \when{X_{\mathtt{lat}} > 34.89}\)}
        \label{subfig:california_ram_2}
    \end{subfigure}
    \caption{Comparison of different predictions for housing prices in California based on the longitude.
    Subfigure (a) showcases the generalized additive model (GAM),
        while subfigures (b) and (c) demonstrate the RAM components for different latitude ranges:
        \(f(X_{\mathtt{long}}) \when{X_{\mathtt{lat}} \leq 34.89}\) and
        \(f(X_{\mathtt{long}}) \when{X_{\mathtt{lat}} > 34.89}\), respectively.
        We observe, that although the EBM model is able to capture the overall trend in the data,
        it also exhibits a large amount of variance.}
    \label{fig:california_housing}
\end{figure}

@@

%\section{Appendix}
%
%\subsection{Regional DALE formulation}
%\label{sec:regional_dale}
%
%Below we provide the regional DALE formulation for the bin-effect, the bin-deviation and the aggregated deviation on a subregion \(\mathcal{R}_{st}\):
%
%\begin{equation}
%    \label{eq:bin_effect_subregion}
%    \hat{\mu}_{\mathcal{R}_{st}}(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k \cap \mathcal{R}_{st}|} \sum_{i:\mathbf{x}^i \in \mathcal{S}_k \cap \mathcal{R}_{st} } \dfdx(\mathbf{x}^i)
%\end{equation}
%
%\begin{equation}
%  \label{eq:bin_deviation_subregion}
%  \hat{\sigma}_{\mathcal{R}_{st}}^2(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k \cap \mathcal{R}_{st}| - 1}
%\sum_{i:\mathbf{x}^i \in \mathcal{S}_k \cap \mathcal{R}_{st} } \left ( \dfdx(\mathbf{x}^i) -
%    \hat{\mu}(z_{k-1}, z_k) \right )^2
%\end{equation}
%
%\begin{equation}
%  \label{eq:SE}
%  \mathcal{H}_{\mathcal{R}_{sk}} = \sum_{k=1}^K (z_k - z_{k-1})^2 \hat{\sigma}_{\mathcal{R}_{st}}^2(z_{k-1}, z_k)
%\end{equation}

%\subsection{Algorithmic Details of Subregion Detection}
%
%\begin{algorithm}
%\caption{SplitDataset}
%\label{alg:split_dataset}
%\SetAlgoLined
%\SetKwInOut{Input}{Input}
%\SetKwInOut{Output}{Output}
%\Input{X\_list, J\_list, c, val, is\_cat}
%\Output{X\_split, J\_split}
%X\_split $\gets$ []\;
%J\_split $\gets$ []\;
%\For{i = 1 to len(X\_list)}{
%    X= X\_list[i]\;
%    J= J\_list[i]\;
%    \uIf{is\_cat}{
%        ind\_1 $\gets$ X[:, c] $=$ val\;
%        ind\_2 $\gets$ X[:, c] $\neq$ val\;
%    }
%    \Else{
%        ind\_1 $\gets$ X[:, c] $\leq$ val\;
%        ind\_2 $\gets$ X[:, c] $>$ val\;
%    }
%    Append X[ind\_1], X[ind\_2] to X\_split\;
%    Append J[ind\_1], J[ind\_2] to J\_split\;
%}
%\Return X\_split, J\_split
%\end{algorithm}
